---
title: "DataCamp_Intermediate-Data-Visualization-with-ggplot2"
output: html_notebook
author: "Andrex Ibiza, MBA"
date: 2025-01-31
---

# Chapter 1: Statistics
A picture paints a thousand words, which is why R ggplot2 is such a powerful tool for graphical data analysis. In this chapter, you’ll progress from simply plotting data to applying a variety of statistical methods. These include a variety of linear models, descriptive and inferential statistics (mean, standard deviation and confidence intervals) and custom functions.

## Video 1.1: Stats with geoms

### ggplot2, course 2
- Statistics
- Coordinates
- Facets
- Data Visualization Best Practices

### Statistics layer
- Two categories of functions:
    - Called from within a geom
    - Called independently
- `stats_`

### `geom_` <-> `stat_`
```{r}
p <- ggplot(iris, aes(x = Sepal.Width))
p + geom_histogram()
```

```{r}
p <- ggplot(iris, aes(x = Sepal.Width))
p + geom_histogram()
p + geom_bar()

```
```{r}
p <- ggplot(mtcars, aes(x = factor(cyl),  fill = factor(am)))
p + geom_bar()
p + stat_count()
```
```markdown
| `stat_` | `geom_` |
| --- | --- |
| `stat_bin` | `geom_histogram()`, `geom_freqpoly() |
| `stat_count()` | `geom_bar()` |
```

### `stat_smooth()`
```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  stat_smooth()

# geom_smooth() using method = 'loess' and formula 'y ~ x'
```
### `stat_smooth(se = FALSE)`
```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  stat_smooth(se = FALSE)

# geom_smooth() using method = 'loess' and formula 'y ~ x'
```

### `geom_smooth(span = 0.4)`
```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  stat_smooth(se = FALSE, span = 0.4)

# geom_smooth() using method = 'loess' and formula 'y ~ x'
```

### `geom_smooth(method = "lm")`
```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE)
```

### `geom_smooth(fullrange = TRUE)`
```{r}
ggplot(iris, aes(x = Sepal.Length,
                 y = Sepal.Width,
                 color = Species)) +
  geom_point() +
  geom_smooth(method = "lm", 
              fullrange = TRUE)
```
```markdown
| `stat_` | `geom_` |
| --- | --- |
| `stat_bin` | `geom_histogram()`, `geom_freqpoly() |
| `stat_count()` | `geom_bar()` |
| `stat_smooth()` | `geom_smooth()` |
```
### Other `stat_` functions
```markdown
| `stat_` | `geom_` |
| --- | --- |
| `stat_boxplot()` | `geom_boxplot()` |
| `stat_bindot()` | `geom_dotplot()` |
| `stat_bin2d()` | `geom_bin2d()` |
| `stat_binhex()` | `geom_hex()` |
| `stat_contour()` | `geom_contour()` |
| `stat_quantile()` | `geom_quantile()` |
| `stat_sum()` | `geom_count()` |
```

## Video 1.2: Stats: sum and quantile

### Recall from course 1
```markdown
| | Cause of Over-plotting | Solutions |
| --- | --- | --- |
| 1 | Large datasets | Alpha-blending, hollow circles, and point size |
| 2 | Aligned values on a single axis | As above, plus change position |
| 3 | Low-precision data | Position: jitter |
| 4 | Integer data | Position: jitter |
```

### Plot counts to overcome over-plotting
```markdown
| | Cause of Over-plotting | Solutions | Here |
| --- | --- | --- | --- |
| 1 | Large datasets | Alpha-blending, hollow circles, and point size |
| 2 | Aligned values on a single axis | As above, plus change position |
| 3 | Low-precision data | Position: jitter | `geom_count()` |
| 4 | Integer data | Position: jitter | `geom_count()` |
```

### Low precision (& integer) data
```{r}
p <- ggplot(iris, aes(Sepal.Length, Sepal.Width))
p + geom_point() + ggtitle("Low precision (& integer) data")
```

### Jittering may give a wrong impression
```{r}
p + geom_jitter(alpha = 0.5,
                width = 0.1,
                height = 0.1) +
  ggtitle("Jittering may give a wrong impression")
```
### `geom_count()`
```{r}
p + geom_count() + ggtitle("geom_count()")
```





