---
title: "Introduction to the Tidyverse"
language: "R"
publisher: "DataCamp"
instructor: [David Robinson](https://www.datacamp.com/instructors/drobinson)
notes_author: [Andrex Ibiza, MBA](andrexibiza@gmail.com)
course-link: https://app.datacamp.com/learn/courses/introduction-to-the-tidyverse
certificate-of-accomplishment: https://www.datacamp.com/completed/statement-of-accomplishment/course/463b056a5a581d5cfe98371bb1d274b9f2d83300
date_completed: 2023-09-09
last_updated: 2025-02-24
---

# Chapter 1: Data Wrangling

## Video 1.1: The `gapminder` dataset

### Loading packages

```{r}
library(gapminder)
library(dplyr)

gapminder
```

## Video 1.2: The `filter` verb

### Filtering for one year

```{r}
gapminder %>% 
  filter(year == 2007)
```

```{r}
gapminder %>% 
  filter(country == "United States")
```

### Filtering for two variables

```{r}
gapminder %>% 
  filter(year == 2007, country == "United States")
```

## Video 1.3: The `arrange` verb

### Sorting with `arrange`

```{r}
gapminder %>% 
  arrange(gdpPercap)
```

### Sorting in descending order

```{r}
gapminder %>% 
  arrange(desc(gdpPercap))
```

### Filtering then arranging

```{r}
gapminder %>% 
  filter(year == 2007) %>% 
  arrange(desc(gdpPercap))
```

## Video 1.4: The `mutate` verb

### Using `mutate` to change a variable

```{r}
gapminder %>% 
  mutate(pop = pop / 1000000)
```

### Using `mutate` to add a new variable

```{r}
gapminder %>% 
  mutate(gdp = gdpPercap * pop)
```

### Combining verbs

```{r}
gapminder %>% 
  mutate(gdp = gdpPercap * pop) %>% 
  filter(year == 2007) %>% 
  arrange(desc(gdp))
```

# Chapter 2: Data Visualization

## Video 2.1: Visualizing with `ggplot2`

### Variable assignment

```{r}
gapminder_2007 <- gapminder %>% 
  filter(year == 2007)
gapminder_2007
```

### Visualizing with `ggplot2`

```{r}
library(ggplot2)
library(ggthemes)
ggplot(gapminder_2007, aes(
                            x = gdpPercap,
                            y = lifeExp
                          )
       ) + geom_point() +
           ggtitle("Life Expectancy vs. GDP per Capita") +
           xlab("GDP per Capita") + 
           ylab("Life Expectancy")
```

## Video 2.2: Log scales

### Scatterplot

```{r}
library(ggplot2)
ggplot(gapminder_2007, aes(x = gdpPercap,
                      y = lifeExp
                      )
) + geom_point()
```

### Log scale

```{r}
ggplot(gapminder_2007, aes(x = gdpPercap,
                      y = lifeExp)
       ) + geom_point() +
           scale_x_log10()
```

### Video 2.3: The `color` aesthetic

```{r}
ggplot(gapminder_2007, aes(x = gdpPercap,
                      y = lifeExp,
                      color = continent)
       ) + geom_point() +
           scale_x_log10() +
           geom_jitter()
```

### The `size` aesthetic

```{r}
ggplot(gapminder_2007, aes(x = gdpPercap,
                      y = lifeExp,
                      color = continent,
                      size = pop)
       ) + geom_point() +
           scale_x_log10() +
           geom_jitter() +
           theme_minimal()
```

## Video 2.3: Faceting

```{r}
ggplot(gapminder_2007, aes(x = gdpPercap,
                           y = lifeExp)) +
  geom_point() +
  scale_x_log10() +
  facet_wrap(~ continent)
```

See notes from Intermediate Data Visualization for more on faceting.

# Chapter 3: Grouping and Summarizing

## Video 3.1: The `summarize` verb

### Extracting data

```{r}
gapminder %>% 
  filter(country == "United States", year == 2007)
```

### The `summarize` verb

```{r}
gapminder %>% 
  summarize(meanLifeExp = mean(lifeExp))
```

### Summarizing one year

```{r}
gapminder %>% 
  filter(year == 2007) %>% 
  summarize(meanLifeExp = mean(lifeExp))
```

### Summarizing into multiple columns

```{r}
gapminder %>% 
  filter(year == 2007) %>% 
  summarize(meanLifeExp = mean(lifeExp),
            totalPop = sum(pop))
```

### Functions you can use for summarizing

-   `mean`

-   `sum`

-   `median`

-   `min`

-   `max`

## Video 3.2: The `group_by` verb

### The `summarize` verb

```{r}
gapminder %>% 
  filter(year == 2007) %>% 
  summarize(meanLifeExp = mean(lifeExp),
            totalPop = sum(pop))
```

### Summarizing by year

```{r}
gapminder %>% 
  group_by(year) %>% 
  summarize(meanLifeExp = mean(lifeExp),
            totalPop = sum(pop)
            )
```

### Summarizing by continent

```{r}
gapminder %>% 
  filter(year == 2007) %>% 
  group_by(continent) %>% 
  summarize(meanLifeExp = mean(lifeExp),
            totalPop = sum(pop))
```

### Summarizing by continent and year

```{r}
gapminder %>% 
  group_by(year, continent) %>% 
  summarize(totalPop = sum(pop),
            meanLifeExp = mean(lifeExp)
            )
```

## Video 3.3: Visualizing Summarized Data

### Summarizing by year

```{r}
by_year <- gapminder %>% 
  group_by(year) %>% 
  summarize(totalPop = sum(pop),
            meanLifeExp = mean(lifeExp)
            )
by_year
```

### Visualizing population over time

```{r}
ggplot(by_year, aes(x = year,
                    y = totalPop)
       ) +
  geom_point()
```

### Starting axis at zero

```{r}
ggplot(by_year, aes(x = year,
                    y = totalPop)
       ) + 
  geom_point() +
  expand_limits(y = 0)
```

### Summarizing by year and continent

```{r}
by_year_continent <- gapminder %>% 
  group_by(year, continent) %>% 
  summarize(totalPop = sum(pop),
            meanLifeExp = mean(lifeExp))
by_year_continent
```

### Visualizing population by year and continent

```{r}
ggplot(by_year_continent, aes(x = year,
                              y = totalPop,
                              color = continent)
       ) + geom_point() +
           expand_limits(y = 0)
```

# Chapter 4: Types of Visualizations

## Video 4.1: Line Plots

```{r}
ggplot(by_year_continent, aes(x = year,
                           y = meanLifeExp,
                           color = continent)) +
  geom_line() +
  expand_limits(y = 0)
```

## Video 4.2: Bar Plots

```{r}
by_continent <- gapminder %>% 
  filter(year == 2007) %>% 
  group_by(continent) %>% 
  summarize(meanLifeExp = mean(lifeExp))

by_continent
```

### Bar plot

```{r}
ggplot(by_continent, aes(x = continent,
                         y = meanLifeExp)) +
  geom_col()
```

## Video 4.3: Histograms

```{r}
ggplot(gapminder_2007, aes(x = lifeExp)) +
  geom_histogram()
```

### Adjusting bin width

```{r}
ggplot(gapminder_2007, aes(x = lifeExp)) +
  geom_histogram(binwidth = 5)
```

### `scale_x_log10()`

```{r}
ggplot(gapminder_2007, aes(x = lifeExp)) +
  geom_histogram(binwidth = 5)

```

## Video 4.4: Box plots

```{r}
ggplot(gapminder_2007, aes(x = continent,
                           y = lifeExp)) +
  geom_boxplot()
```
