---
title: "Titanic Survival Analysis with Neural Networks"
output: html_notebook
---

# Introduction
This script represents my continued exploration of the Titanic dataset, aiming to enhance the predictive accuracy of survival outcomes for passengers aboard the ill-fated ship. Building upon my initial model, which achieved approximately 70% accuracy, this iteration seeks to delve deeper into the dataset by employing a more sophisticated approach to data exploration, handling missing values, and engineering insightful features. By leveraging advanced data manipulation techniques and machine learning models, this project aspires to uncover hidden patterns and improve the robustness of predictions. The ultimate goal is to surpass previous performance benchmarks and gain a more comprehensive understanding of the factors influencing survival, as part of the ongoing Kaggle Titanic competition.

## Files
* `gender_submission.csv`: example of what the final submitted file should look like with two columns: `PassengerID` and `Survived`.
* `train.csv`: labeled data (`Survived`) used to build the model. 11 columns
* `test.csv`: 12 columns

## Data dictionary

| Variable	| Definition | Key | Notes |
| --- | --- | --- | --- |
| survival	| Survival	| 0 = No, 1 = Yes | --- |
| pclass	| Ticket class	| 1 = 1st, 2 = 2nd, 3 = 3rd | Proxy for SES- 1st=upper, 2nd=middle, 3rd=lower |
| sex	| Sex | --- | --- |
| Age	| Age in years | --- | Age is fractional if less than 1. If the age is estimated, is it in the form of xx.5 |
| sibsp	| # of siblings / spouses aboard the Titanic | --- | Sibling = brother, sister, stepbrother, stepsister; Spouse = husband, wife (mistresses and fiancés were ignored) |
| parch	| # of parents / children aboard the Titanic | --- | Parent = mother/father, Spouse = husband, wife (mistresses and fiances ignored). Some children travelled only with a nanny, therefore parch=0 for them. |
| ticket | Ticket number | --- | --- |
| fare	| Passenger fare | --- | --- |
| cabin	| Cabin number | --- | --- |
| embarked | Port of Embarkation | C = Cherbourg, Q = Queenstown, S = Southampton | --- ||mpton | --- |

```{r}
# Load packages
library(caret)        # machine learning
library(dplyr)        # data manipulation
library(ggplot2)      # viz
library(Hmisc)        # robust describe() function
library(naniar)       # working with missing data
library(randomForest) # inference model

# Load train and test data
train <- read.csv("train.csv", stringsAsFactors = FALSE)
test <- read.csv("test.csv", stringsAsFactors = FALSE)
head(train) #--loaded successfully
head(test)  #--loaded successfully

# Evaluate structure and data types
# str(train)
# str(test)
# 
# describe(train)
# train has missing values: Age 177, Cabin 687, Embarked 2
# describe(test)
# test has missing values: Cabin 327, Fare 1, Age 86
```

```{r}
# DATA CLEANING AND PREPROCESSING
# 1) Encode categorical variables
# [X] Encode Sex as numeric factor
train$Sex <- as.numeric(ifelse(train$Sex == "male", 1, 0)) # v2.2 added as.factor() to coerce output
test$Sex <- as.numeric(ifelse(test$Sex == "male", 1, 0))
head(train[, "Sex"]) #--encoded successfully
head(test[, "Sex"]) #--encoded successfully
```


```{r}
# [X] Convert Pclass to an ordinal factor
train$Pclass <- as.numeric(factor(train$Pclass, levels = c(1, 2, 3), ordered = TRUE))
test$Pclass <- as.numeric(factor(test$Pclass, levels = c(1, 2, 3), ordered = TRUE))
head(train[, "Pclass"]) #--encoded successfully
head(test[, "Pclass"]) #--encoded successfully
```

```{r}
# [X] One-hot encode Embarked
embarked_train_one_hot <- model.matrix(~ Embarked - 1, data = train)
embarked_test_one_hot <- model.matrix(~ Embarked - 1, data = test)

# Add the one-hot encoded columns back to the dataset
train <- cbind(train, embarked_train_one_hot)
test <- cbind(test, embarked_test_one_hot)

# Verify encoding:
#head(train[, c("Embarked", "EmbarkedC", "EmbarkedQ", "EmbarkedS")])
#head(test[, c("Embarked", "EmbarkedC", "EmbarkedQ", "EmbarkedS")])

# -- looks perfect, let's not forget about imputing our 2 missing values
# Impute 2 missing Embarked values with the mode
train$Embarked[train$Embarked == ""] <- NA
embarked_mode <- names(sort(table(train$Embarked), decreasing = TRUE))
train$Embarked[is.na(train$Embarked)] <- embarked_mode

# verify imputation
describe(train$Embarked)
```

```{r}
##v2.2 also want to explicitly cast the values in EmbarkedC, EmbarkedQ, and EmbarkedS as factors.
train$EmbarkedC <- as.numeric(as.factor(train$EmbarkedC))
test$EmbarkedC <- as.numeric(as.factor(test$EmbarkedC))
train$EmbarkedQ <- as.numeric(as.factor(train$EmbarkedQ))
test$EmbarkedQ <- as.numeric(as.factor(test$EmbarkedQ))
train$EmbarkedS <- as.numeric(as.factor(train$EmbarkedS))
test$EmbarkedS <- as.numeric(as.factor(test$EmbarkedS))

## SibSp and Parch should be integers
train$SibSp <- as.numeric(train$SibSp)
test$SibSp <- as.numeric(test$SibSp)
train$Parch <- as.numeric(train$Parch)
test$Parch <- as.numeric(test$Parch)

# Survived needs to be a factor
train$Survived <- as.numeric(as.factor(train$Survived))
```

```{r}
# 3) Address missing values
# Age - Train
#--Predict missing ages using other features
train_age_data <- train %>% 
    select(Age, Pclass, Sex, SibSp, Parch, Fare, EmbarkedC, EmbarkedQ, EmbarkedS)

# head(train[, c("Age", "Pclass", "Sex", "SibSp", "Parch", "Fare", "EmbarkedC", "EmbarkedQ", "EmbarkedS")])
#--verified that all these columns are formatted properly

train_age_complete <- train_age_data %>% filter(!is.na(Age))
train_age_missing <- train_age_data %>% filter(is.na(Age))

set.seed(666)
cv_control <- trainControl(method = "cv", number = 10) #v2.2 10-fold cross-validation for imputing missing ages
train_age_cv_model <- train(
  Age ~ Pclass + Sex + SibSp + Parch + Fare + EmbarkedC + EmbarkedQ + EmbarkedS,
  data = train_age_complete,
  method = "rf",
  trControl = cv_control,
  tuneLength = 3
)
print(train_age_cv_model)
```

```{r}
# Use the best model to predict missing ages
predicted_train_ages <- predict(train_age_cv_model, newdata = train_age_missing)

# Impute the predicted ages back into the train dataset
train$Age[is.na(train$Age)] <- predicted_train_ages
describe(train$Age)
```


```{r}
#--Age in test data
# Preprocess the test data for Age imputation
test_age_data <- test %>% 
  select(Age, Pclass, Sex, SibSp, Parch, Fare, EmbarkedC, EmbarkedQ, EmbarkedS)

test_age_missing <- test_age_data %>% filter(is.na(Age))
test_age_complete <- test_age_data %>% filter(!is.na(Age))

# Use the trained train_age_cv_model to predict missing ages in the test dataset
predicted_test_ages <- predict(train_age_cv_model, newdata = test_age_missing)

# Impute the predicted ages back into the test dataset
test$Age[is.na(test$Age)] <- predicted_test_ages

n_miss(test$Age)
```

```{r}
library(stringr)

## Feature Engineering - transform Name into Title
# Update the regex pattern to include all titles
title_pattern <- "Mr|Mrs|Miss|Master|Don|Rev|Dr|Mme|Ms|Major|Lady|Sir|Mlle|Col|Capt|Countess|Jonkheer"

# Extract titles using the regex title_pattern
train$Title <- as.factor(str_extract(train$Name, title_pattern))
test$Title <- as.factor(str_extract(test$Name, title_pattern))

str(train)
str(test)
```

```{r}
# Convert empty strings to NA in Cabin
train$Cabin[train$Cabin == ""] <- NA
test$Cabin[test$Cabin == ""] <- NA

# Create new `Deck` feature
train$Deck <- as.factor(ifelse(!is.na(train$Cabin), substr(train$Cabin, 1, 1), "U"))
test$Deck <- as.factor(ifelse(!is.na(test$Cabin), substr(test$Cabin, 1, 1), "U"))

# Verify the new Cabin and Deck features
head(train[, c("Cabin", "Deck")])
head(test[, c("Cabin", "Deck")])
```

```{r}
# Encode the HasCabin variable:
train$HasCabin <- as.numeric(as.factor(ifelse(!is.na(train$Cabin), 1, 0)))
test$HasCabin <- as.numeric(as.factor(ifelse(!is.na(test$Cabin), 1, 0)))

# describe(train$HasCabin) # - perfect
head(train[, c("Cabin", "HasCabin")])  #looks good
head(test[, c("Cabin", "HasCabin")]) 

n_miss(train$HasCabin)
n_miss(test$HasCabin)
```

```{r}
# Create the FamilySize feature
train$FamilySize <- as.numeric(train$SibSp + train$Parch + 1)
test$FamilySize <- as.numeric(test$SibSp + test$Parch + 1)

# Inspect the new feature
head(train[, "FamilySize"])
head(test[, "FamilySize"])

# describe(train)
# describe(test)
#--test still has 1 missing fare - impute with the median
test$Fare[is.na(test$Fare)] <- median(test$Fare, na.rm = TRUE)
describe(test)
```

```{r}
# drop `Embarked` here because it was causing a duplicate column error
train <- train %>% select(-Embarked)
test <- test %>% select(-Embarked)

# Create the GroupSize feature based on Ticket
train$GroupSize <- train %>%
  group_by(Ticket) %>%
  mutate(GroupSize = n()) %>%
  ungroup() %>%
  pull(GroupSize)

test$GroupSize <- test %>%
  group_by(Ticket) %>%
  mutate(GroupSize = n()) %>%
  ungroup() %>%
  pull(GroupSize)

train$GroupSize <- as.numeric(train$GroupSize)
test$GroupSize <- as.numeric(test$GroupSize)

# Inspect the new feature
head(train[, c("Ticket", "GroupSize")])
head(test[, c("Ticket", "GroupSize")])
```

```{r}
# Create the FarePerPerson feature
train$FarePerPerson <- as.numeric(train$Fare / train$GroupSize)
test$FarePerPerson <- as.numeric(test$Fare / test$GroupSize)

# Inspect the new feature
head(train[, c("Ticket", "Fare", "GroupSize", "FarePerPerson")])
head(test[, c("Ticket", "Fare", "GroupSize", "FarePerPerson")])
```

```{r}
# Apply log transformation to Fare and FarePerPerson
#--plot shape before transformation?
ggplot(train, aes(x = Fare)) +
  geom_histogram(bins=20) +
  theme_minimal() +
  ggtitle("Fare (before transforming)")

#--note an extreme outlier over 500!
train$Fare <- log(train$Fare + 1)
train$FarePerPerson <- log(train$FarePerPerson + 1)
test$Fare <- log(test$Fare + 1)
test$FarePerPerson <- log(test$FarePerPerson + 1)
head(train[, c("Fare", "FarePerPerson")])
head(test[, c("Fare", "FarePerPerson")])

ggplot(train, aes(x = Fare)) +
  geom_histogram(bins=20) +
  theme_minimal() +
  ggtitle("Log Transformed Fare")
```

```{r}
ggplot(train, aes(x = FarePerPerson)) +
  geom_histogram(bins=20) +
  theme_minimal() +
  ggtitle("Log Transformed FarePerPerson")
```

```{r}
str(train)
str(test)
```

```{r}
# Create the ChildInFamily feature
train$ChildInFamily <- as.numeric(as.factor(ifelse(train$Age < 15 & train$FamilySize > 1, 1, 0)))
test$ChildInFamily <- as.numeric(as.factor(ifelse(test$Age < 15 & test$FamilySize > 1, 1, 0)))

# Inspect the new feature
head(train[, c("Age", "FamilySize", "ChildInFamily")])
head(test[, c("Age", "FamilySize", "ChildInFamily")])
```
```{r}
# Fixing Data Type Issues in Titanic Keras Model

# Ensure Survived is properly formatted
train$Survived <- as.factor(train$Survived)
y_train <- as.numeric(as.character(train$Survived))

# Ensure all categorical features are converted to numeric
train$Pclass <- as.numeric(train$Pclass)
test$Pclass <- as.numeric(test$Pclass)

train$Sex <- as.numeric(train$Sex)
test$Sex <- as.numeric(test$Sex)

train$EmbarkedC <- as.numeric(train$EmbarkedC)
test$EmbarkedC <- as.numeric(test$EmbarkedC)

train$EmbarkedQ <- as.numeric(train$EmbarkedQ)
test$EmbarkedQ <- as.numeric(test$EmbarkedQ)

train$EmbarkedS <- as.numeric(train$EmbarkedS)
test$EmbarkedS <- as.numeric(test$EmbarkedS)

# Fix HasCabin encoding
train$HasCabin <- as.numeric(!is.na(train$Cabin))
test$HasCabin <- as.numeric(!is.na(test$Cabin))

# Normalize Fare and Age
train$Fare <- scale(train$Fare)
test$Fare <- scale(test$Fare)

train$Age <- scale(train$Age)
test$Age <- scale(test$Age)

train$FarePerPerson <- scale(train$FarePerPerson)
test$FarePerPerson <- scale(test$FarePerPerson)

# Drop unnecessary columns
train <- train %>% select(-Name, -Ticket, -Cabin)
test <- test %>% select(-Name, -Ticket, -Cabin)

# Convert features to numeric matrix
x_train <- as.matrix(sapply(train %>% select(-Survived), as.numeric))
x_test <- as.matrix(sapply(test, as.numeric))

y_train <- as.numeric(train$Survived)

# Ensure no missing values
if (any(is.na(x_train)) || any(is.na(y_train))) {
  stop("Missing values detected in x_train or y_train. Please clean the data.")
}

# Ensure x_train is a proper numeric matrix
if (!is.matrix(x_train)) {
  stop("x_train is not a matrix. Check feature selection and conversions.")
}

if (!all(sapply(x_train, is.numeric))) {
  stop("Non-numeric columns detected in x_train. Check data preprocessing.")
}
  stop("Missing values detected in training data. Please handle missing values before training.")
}

# Check structure before training
str(x_train)
str(y_train)
```

```{r}
# Load necessary libraries
library(keras)

# Convert features to numeric matrices:
# For training data, drop the 'Survived' column (which is the target)
x_train <- as.matrix(sapply(train[, setdiff(names(train), "Survived")], as.numeric))
y_train <- as.numeric(as.character(train$Survived))

# For test data, use all columns as features (assuming test does not have 'Survived')
x_test <- as.matrix(sapply(test, as.numeric))

# (Optional) Check dimensions to verify they match expected input sizes
cat("x_train shape:", dim(x_train), "\n")
cat("x_test shape:", dim(x_test), "\n")

# Define the model using the Functional API (explicit Input layer)
inputs <- layer_input(shape = c(ncol(x_train)))
outputs <- inputs %>%
  layer_dense(units = 64, activation = "relu") %>%
  layer_dropout(rate = 0.4) %>%
  layer_dense(units = 32, activation = "relu") %>%
  layer_dropout(rate = 0.3) %>%
  layer_dense(units = 1, activation = "sigmoid")  # Using 'sigmoid' for binary classification

model <- keras_model(inputs = inputs, outputs = outputs)

# Compile the model
model$compile(
  optimizer = optimizer_adam(),
  loss = "binary_crossentropy",  # Binary crossentropy for classification
  metrics = list("accuracy")
)

# Train the model
history <- model$fit(
  x_train, y_train,
  epochs = 30,
  batch_size = 32,
  validation_split = 0.2
)

# Predict on the test set
predictions <- model$predict(x_test)

# Convert predictions to binary (0/1) values using a 0.5 threshold and update the test dataset
test$Survived <- ifelse(predictions > 0.5, 1, 0)

# Check the first 40 predictions
head(test$Survived, 40)
```