---
title: "Working with Dates and Times in R"
output: html_notebook
---

# Chapter 1: Dates and Times in R

## Video 1.1: Introduction to dates

### Dates
Different conventions in different places

27th Feb 2013
* NZ: 27/2/2013
* USA: 2/27/2013

### ISO 8601 YYYY-MM-DD
* Values ordered from the largest to smallest unit of time.
* Each has a fixed number of digits, must be padded with leading zeros.
* Either, no separators for computers, or `-` in dates.
* 1st of January 2011 -> 2011-01-01

### Dates in R
* Packages that import dates: `readr`, `anytime`.

```{r}
2003-02-27
# interpreted as mathematical function
```

```{r}
"2003-02-27"
# interpreted as character string
```

```{r}
str("2003-02-27")
# interpreted as character string
```
```{r}
as.Date("2003-02-27")
```

```{r}
str(as.Date("2003-02-27"))
```

### Specifying dates
```{r}
# The date R 3.0.0 was released
x <- "2013-04-03"

# Examine structure of x
str(x)
```

```{r}
# Use as.Date() to interpret x as a date
x_date <- as.Date(x)

# Examine structure of x_date
str(x_date)

# Store April 10 2014 as a Date
april_10_2014 <- as.Date("2014-04-10")
```

### Automatic import
Sometimes you'll need to input a couple of dates by hand using `as.Date()` but it's much more common to have a column of dates in a data file.

Some functions that read in data will automatically recognize and parse dates in a variety of formats. In particular the import functions, like `read_csv()`, in the readr package will recognize dates in a few common formats.

There is also the `anytime()` function in the `anytime` package whose sole goal is to automatically parse strings as dates regardless of the format.

##


